Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-01-19 manu_draft.Rmd
started incorporating JH edits, intro done

Accurate and flexible trend assessment methods are valuable tools for describing historical changes in environmental monitoring datasets.  However, mismatches between time scales present in the monitoring data and scales of analysis or management questions can create challenges for accurately assessing trends with full propagation of uncertainty across methods. Here, we propose a novel application of Generalized Additive Models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in South San Francisco Estuary.  GAMs are a general method for modelling a response variable as the additive sum of one or more smoother splines fit to predictor variables.  These models have shown promise in other applications to water quality monitoring data to separate long-term (i.e., annual or decadal) trends from seasonal variation.  Our proposed methods address the specific goal of estimating seasonal averages in a response variable and using those averages plus uncertainty to assess inter-annual trends, neither of which are addressed by previous applications of GAMs to environmental monitoring data.  The approach first demonstrates that nearly identical descriptions of chlorophyll as a function of time can be obtained using different smoothing splines for annual or seasonal components of the time series.  We then extract seasonal averages for an a priori period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.  Finally, we demonstrate how trends can be assessed with mixed-meta regression analyses that evaluates changes in the seasonal averages between years with full propagation of uncertainty in the estimates.  Overall, this approach leverages the ability of GAMs to identify an optimal amount of smoothing from the raw data and uses those results to accurately assess trends that minimize the likelihood of false positive or negative results, which is particularly important for time series with missing observations or varying sample effort over time.

@@

Many environmental monitoring programs generate fine-scale but potentially irregular time-series data to assess long-term trends for regulatory, management, or research purposes. The mismatch between the scales of monitoring versus analysis questions or management goals can lead to statistical challenges [@Urquhart98;@Cumming06;@Forbes18]. At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events, management, or seasonal changes).  Such fluctuations may not be of direct interest for longer-term trends or may not be well-suited to multi-scale smoothing methods.  However, aggregate features of seasonal patterns that integrate or summarize short-term fluctuations may be of interest.  In this paper, we describe methods to estimate across-year trends of within-year features of interest such as a seasonal average, seasonal peak, or seasonal timing of events, while accounting for uncertainties across analysis steps.

Previous methods for water quality trend analysis can be generalized into four basic approaches.  Seasonal Kendall tests or related non-parametric approaches have been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years [@Hirsch82;@Helsel20].  A literature survey of @Wan17 revealed non-parametric approaches to be the most commonly used methods in long-term water quality trend analysis, yet they have limited scope.  For descriptive decomposition of long-term monitoring data, they do not account for changes occurring at different temporal scales, do not adequately evaluate irregularly spaced data, do not incorporate other explanatory variables, and do not estimate a model that could be useful for other purposes.  Thus, while these non-parametric approaches have some degree of robustness to assess magnitude and direction of trends, they apply only to narrow goals.

The seasonal trend decomposition using loess (STL) decomposes a time series into additive components of a long-term trend, a seasonal pattern, and residuals [@Cleveland90;@Cloern10;@Stow15].  While useful and widely applied, this method also has important limitations.  STL decomposition does not incorporate explanatory variables, it is defined more as an algorithm of statistical steps than as a coherent statistical model [e.g., @Wan17], and it does not usually include standard errors to allow hypothesis testing [but see @Hafen10].  Conventional STL approaches may also over-simplify trends into absolute components that do not change over time, e.g., a seasonal estimate that is constant across years.  This limitation presents challenges when addressing questions relevant to long-term water quality data, such as timing of seasonal peaks that can suggest system response to changing  environmental conditions [@Cloern10;@Navarro12].

The more recently developed method of weighted regression on time, discharge, and season (WRTDS) uses a more general local regression scheme than STL [@Hirsch10;@Beck15;@Beck18b].  Designed for river data where separating the effect of discharge on constituent concentration is important, WRTDS estimates a moving window regression model with components that allows parameters to vary smoothly in relation to both time and discharge.  This yields parameters that are specific to season, year, and flow regime.  Conceptually, the approach is similar to localized multi-polynomial smoothing methods, although the application was developed specifically for describing long-term water quality trends.  Standard error estimates of predictions are available through a "block bootstrap" approach that uses Monte Carlo estimates of false positive rates from the model results [@Hirsch15]. Although a useful addition to the original method in @Hirsch10, the approach requires extensive resampling as a post-hoc application to a previously fitted model.

The final and most recent approach is to use smoothing splines to separate fluctuations on different time scales and do so within the larger framework of generalized additive models (GAMs) [@He06;@Morton08;@Pearce11;@Haraguchi15;@Murphy19].  These may be seen as generalizing the concepts behind STL and WRTDS. In statistics, the evolution of non-parametric regression methods has largely converged on GAMs rather than more generalized kernel smoothing methods used by both STL and WRTDS.  Kernel smoothing and spline smoothing are closely related, and a key challenge for each is to determine the appropriate degree of smoothing.  For example, WRTDS can potentially give results similar to the spline-based smoothing methods described next, although at higher computational expense and with the limitation that uncertainty estimates are not readily obtainable from the original method [@Beck17]. User input is also required to specify an acceptable degree of smoothing used by the "windows" that define the localized fit of WRTDS at each point in the time series.  These windows are conceptually similar to the kernel (or bandwidth) that used in more conventional smoothing methods. There is no simple rule to guide the choice of defining an appropriate size and a tradeoff between over- and under-smoothing is a hallmark of these approaches.

Compared to kernel smoothing methods, GAMs have various advantages.  They are formulated using "basis functions", and these can be customized for needs such as cyclic splines (e.g., for an annual pattern) and low-dimensional interactions.  They can naturally include both parametric (e.g. linear or quadratic) components and non-parametric (spline) components.  Importantly, multiple approaches to automatically determine the optimal degree of smoothness have been developed, based on likelihood and/or optimizing out-of-sample prediction error.  They have natural frequentist and Bayesian interpretations, are naturally extensible to include random effects (GAMMs), and have computationally efficient implementations that can be optimized more quickly than other approaches [@Wood17].  For these reasons, GAMs are widely used for non-parametric regression smoothing in many fields.

GAMs have recently been applied to trend analysis of water quality time-series, particularly from long-term monitoring programs [@Haraguchi15;@Murphy19], but with different formulations and goals than given here. For example, in the US, the Chesapeake Bay Program uses GAMs to decompose time-series into long-term and seasonal trends [@Murphy19] and test trend hypotheses between two points in time.  That is related to the methods here, but this paper gives more general methods for analyzing trends of seasonal spline features, describes the relationships among alternative spline formulations when they are used as designed [@Wood03;@Wood17] rather than for ad hoc separation of time scales, and prioritizes full incorporation of uncertainty.  Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and different time lags in describing relationships between response and predictor variables [@Lefcheck17;@Testa18].

Our motivating problem has several needs that are not satisfied by previous methods, but can use GAMs as a starting point. Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as chlorophyll.  However, the seasonal average within each year must account for different sampling times and intervals, and any trend analysis must incorporate the uncertainties in seasonal averages. STL and/or WRTDS could potentially separate seasonal from long-term trends, but doing so is not necessary to determine seasonal averages.  What is needed is an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, allowing for irregular sampling and the non-independence inherent in time-series.  This can be done with GAMs, but we develop an application that is distinct from previous studies. Even if estimates of seasonal averages and their standard errors are available, none of these methods are designed to understand interannual trends in those averages.  A Kendall test would not incorporate the standard errors or reveal useful long-term patterns beyond a significance test. Similarly, STL and WRTDS are not designed for this goal.

We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Estuary, California, USA.  For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.  Analysis of these data is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data may not adequately describe trends.  Examples of long-term trend questions include: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chlorophyll levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or in baseline levels during periods of low productivity?  We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] following signal extraction with GAMs that is new to environmental trend-detection problems.  For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evaluating these trends between years.

>>>>

Accurate and flexible trend assessment methods are valuable tools for describing historical changes in environmental monitoring datasets.  A key requirement is complete propagation of uncertainty through the analysis.  However, this is difficult when there are mismatches between time scales of monitoring data and analysis methods. Here, we propose a novel application of Generalized Additive Models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in South San Francisco Estuary.  GAMs are a general method for modelling a response variable as the additive sum of one or more smoother splines fit to predictor variables.  These models have shown promise in other applications to water quality monitoring data to separate long-term (i.e., annual or decadal) trends from seasonal variation.  Our proposed methods estimate seasonal averages in a response variable with GAMs and then use those ancertainty with mixed-meta regression analyses to quantify inter-annual trends that account for full propagation of error across methods.  We first demonstrate that nearly identical descriptions of temporal changes in l can be obtained using different smoothing splines for annual or seasonal components of the time series.  We then extract seasonal averages for an a priori period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.  Finally, we demonstrate how trends can be quantified with mixed-meta regression analyses that evaluates changes in the seasonal averages between years with full propagation of uncertainty in the estimates.  Overall, this approach leverages the ability of GAMs to identify an optimal amount of smoothing from the raw data and uses those results to accurately assess trends that minimize the likelihood of false positive or negative results, which is particularly important for time series with missing observations or varying sample effort over time.

@@

Accurate quantification of trends must consider variation at different temporal scales regardless of the question of interest, such that ignoring variation at one scale could lead to incorrect conclusions about variation at another scale. Many environmental monitoring programs collect temporally resolved but irregular time-series data to quantify trends for regulatory, management, or research purposes.  The mismatch between the scales of monitoring versus analysis questions or management goals can present statistical challenges [@Urquhart98;@Cumming06;@Forbes18]. At short temporal scales typically less than a year, environmental systems exhibit variability caused by multiple factors (e.g., weather events, management, or seasonal changes).  Such fluctuations may not be of direct interest for inter-annual trends or may not be well-suited to multi-scale smoothing methods.  In this paper, we describe methods to estimate across-year trends of within-year features of interest such as a seasonal average, seasonal peak, or seasonal timing of events, while accounting for uncertainties across analysis steps.

Existing methods that address our objectives in water quality trend analysis can be generalized into four basic approaches: seasonal Kendall tests, seasonal trend decomposition (STL), weighted regression on time, discharge, and season (WRTDS), and generalized additive models (GAMs).  Seasonal Kendall tests or related non-parametric approaches have been used for decades in water quality trend assessments to identify monotonic changes over several years while accounting for the predictable patterns among seasons [@Hirsch82;@Helsel20]. @Wan17 showed that non-parametric approaches have been the most commonly used methods in long-term water quality trend analysis despite critical limitations.  For descriptive decomposition of long-term monitoring data, they assume seasonal patterns within year do not change, require regularly spaced or balanced data, only use time as a descriptor for trends, and do not estimate a model that could be useful for other purposes.  Thus, while these non-parametric approaches have some degree of robustness to assess magnitude and direction of trends, they apply only to narrow goals.

The seasonal trend decomposition using loess (STL) decomposes a time series into additive components of a long-term trend, a seasonal pattern, and residuals [@Cleveland90;@Cloern10;@Stow15].  While useful and widely applied, this method also has important limitations.  STL decomposition does not incorporate explanatory variables besides time, it is defined more as an algorithm of statistical steps than as a coherent statistical model [e.g., @Wan17], and it does not usually estimate standard errors to allow hypothesis testing [but see @Hafen10].  STL methods may also over-simplify trends into absolute components that do not change over time, e.g., a seasonal estimate that is constant across years.  This limitation presents challenges when addressing questions relevant to long-term water quality data, such as timing of seasonal peaks that can suggest system response to changing  environmental conditions [@Cloern10;@Navarro12].

The weighted regression on time, discharge, and season (WRTDS) method addresses the problem of inflexibility in STL by using a more general local regression scheme [@Hirsch10;@Beck15;@Beck18b].  Designed for evaluating water quality in rivers where separating the effect of discharge on constituent concentration is important, WRTDS estimates a moving window regression model with components that allow parameters to vary smoothly in relation to both time and discharge.  This yields parameters that are specific to season, year, and flow regime.  The WRTDS approach is conceptually similar to localized multi-polynomial smoothing methods, with specific application to descriptive variables relevant for water quality constituents (i.e., season, year, and discharge). Standard error estimates of predictions from WRTDS are available through a "block bootstrap" approach that uses Monte Carlo estimates of false positive rates from the model results [@Hirsch15]. Although a useful addition to the original method [@Hirsch10], the approach requires extensive resampling as a post-hoc application to a previously fitted model.  Alternative methods that include standard error estimates simultaneously with model output may be preferred for intensive or more iterative applications

Generalized additive models (GAMs) are central to this paper and form the basis of our fourth and last method to separate fluctuations on different time scales.  GAMs combine one or more smoothing splines to model patterns in data and may be seen as generalizing the concepts behind STL and WRTDS [@He06;@Morton08;@Pearce11;@Haraguchi15;@Murphy19]. In statistics, the evolution of non-parametric regression methods has largely converged on GAMs, rather than kernel smoothing methods used by both STL and WRTDS.  Kernel smoothing and spline smoothing are closely related, and a key challenge for each is to determine the appropriate degree of smoothing.  For example, WRTDS can potentially give results similar to the spline-based smoothing methods used in GAMs, although at higher computational expense and with the limitation that uncertainty estimates are not readily obtainable from the original method [@Beck17]. WRTDS also requires specifying the desired degree of smoothing via the "windows" that define the local weighting at each point in the time series.  The windows in WRTDS are conceptually similar to the kernel (or bandwidth) used in more conventional smoothing methods. A challenge of WRTDS is that there is no clear rule that can be applied to determine the best choice of window size. The associated tradeoff between over- and under-smoothing is a hallmark of these approaches.

GAMs have various advantages compared to kernel smoothing methods and are therefore used for non-parametric regression smoothing in many fields.  The "basis functions" used to formulate GAMs can be customized based on expected patterns in the data. Examples inlcude cyclic splines, which can be used to model seasonal patterns, and low-dimensional interactions.  GAMs have added flexibility because they can include both parametric (e.g. linear or quadratic) components and non-parametric (spline) components.  Multiple approaches have been developed to determine the optimal degree of smoothness.  These approaches are based on likelihood estimates or optimization of out-of-sample prediction error, which address a key concern around methods like WRTDS that do not have analogs for choosing optimal degrees of smoothing.  Further, GAMs have natural frequentist and Bayesian interpretations, are naturally extensible to include random effects (i.e., generalized additive mixed models or GAMMs), and have computationally efficient implementations that can be optimized more quickly than other approaches [@Wood17].

GAMs have been applied previously to evaluate trends in water quality time-series from long-term monitoring programs [@Haraguchi15;@Murphy19]. For example, @Murphy19 used GAMs to decompose water quality time series from Chesapeake Bay into long-term and seasonal trends [@Murphy19] and test trend hypotheses between two points in time.  This study generalizes the approach to analyzing trends of seasonal spline features, describes the relationships among alternative spline formulations when they are used as designed [@Wood03;@Wood17] rather than for ad hoc separation of time scales, and prioritizes full incorporation of uncertainty.  Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and quantifying time lags in relationships between response and predictor variables [@Lefcheck17;@Testa18].

Our motivating problem has several characteristics that are partially addressed by previous methods and can further build on  GAMs as a starting point. Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as chlorophyll.  However, the seasonal average within each year must be robust to inconsistent sampling times and intervals, and any trend analysis must consider the uncertainties in seasonal averages.  The critical need is the ability to obtain an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, even with irregular sampling and serial correlation, which is common to time series data.  This paper develops GAMs as a distinct application that can meet these requirements and simultaneously addressing key limitations of seasonal Kendall tests and the more complex STL and WRTDS methods.  We do so with the explicit goal of quantifying interannual trends in seasonal averages.

We describe and demonstre the proposed methods by analyzing water quality monitoring data from the southern portion of San Francisco Estuary, California, USA.  Approximately twice-monthly monitoring has been conducted for several decades at fixed locations (stations) on the longitudinal axis of the Bay.  Analysis of these data is complicated by irregularities in timing and consistency of data collection which can generate artifacts affecting simple seasonal averages of the data.  We were interested in questions such as: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chlorophyll levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or baseline levels of productivity between blooms?  We provide examples illustrating how these questions can be addressed using methods with GAMs.  We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] after signal extraction with GAMs.  This approach is new to environmental trend-detection problems.  For this step, we provide methods using meta-anlaysis for isolating seasonal trends with reasonalbe certainty from from GAM results and evaluating these trends between years.
